Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-11-12 output/breed_paper.Rnw
Updates with some of Bruce's comments.

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using experimental yield data
over 30+ years at the California Cooperative Rice Research Foundation
Rice Experiment Station. Results show an apparent increase in yield in
side-by-side comparisons between old and new rice varieties when the
yield erosion was not considered. However, the yields of older
varieties eroded at an estimated rate of 29.3 kg ha\tsups{-1}
year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
selection. Once this effect was considered, the yield advantage of
newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
year\tsups{-1}, 90\% credible interval -36.1 to 31.5). This suggests
that side-by-side comparisons reflected the decreasing yields of older
cultivars more than the increasing yield of newer ones. These results
highlight the importance of continuous crop improvement and deployment
of new cultivars as yield will decline over time without it. To
increase the genetic yield potential of rice in the face of this
phenomenon, higher yield targets are needed to counter-act the
decrease over time in the current cultivars.

@@

<<echo=FALSE>>=

library(rstan)
library(rstanarm)
library(lme4)
library(xtable)

@@

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect true or absolute yield
increases over time.

@@

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
Agriculture National Agricultural Statistics Service, 2017). Rapid
gains in grain yield were experienced from approximately 1950 -- 1990,
but annual gains since 1990 have been slower.}
\label{fig:state_yield}
\end{figure}

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the $Rate_{yield\ erosion\ i}$ term.

% Add more detail to the methods section - tell how you get these
% numbers. "Evidence for erosion" - what does this mean for yield
% increases? Erosion 1st, 2nd implications of erosion. 
% Drop off -5 and -6

Data were processed and models were fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval~\citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods (frequentist mixed-effects
model fit by maximum likelihood; See supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

There was general agreement between the estimates from the Bayesian
and classical approaches.  The Bayesian estimates are presented here
due to the inherent flexibility of Bayesian inference, though the
classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
indicates there was no evidence of genetic yield improvements during
the 30+ year study period (Fig. \ref{fig:RES_year}). Individual exceptions to this were
two cultivars (M-204 and M-205) that had higher yields than the
others.

@@

The yields of each cultivar over time were evaluated relative to when
the cultivar was released using the full model which accounts for both
yield erosion and yield improvements over time. The average decrease
in yield over time (i.e., yield erosion) once a
cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
trials (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
has lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed solely
to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:ids}).

@@

When yield erosion is not corrected for in the analysis (i.e.,
omitted from the model), there is estimated to be a small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
full model which corrected for both the year of release and yield
erosion indicates that the estimated change in grain yield per year
in the CCRRF's breeding trials at the RES was \Sexpr{release_yr_coef}
kg ha\tsups{-1} year\tsups{-1}, with both small positive and negative
trends within the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}).

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated by random
intercepts). The marginal estimate for the performance of a cultivar
averaged over the uncertainty in the other parameters in the model
(i.e., the random effect) shows three cultivars with higher than
average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),collapse=',')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).

@@

In California, the cultivars with broad adaptation and adoption over
the 30+ year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in one of these cultivars during the study
period~\citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
and M-402 which is a premium quality cultivar. Other cultivars such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this cultivar is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or
cooler (Espe et al., 2017) temperatures reducing yields. Again, this
is unlikely to be the case in this situation as Espe et al. (2017)
found no significant changes in growing season temperatures (either
high, low or average temps) the Sacramento Valley during the study
period.  Thirdly, yield erosion of a cultivar as has been reported for
the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
and abiotic forces have previously been speculated to be responsible
for yield erosion, since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar may experience
over time are continuously changing~\citep{Peng1998, Peng2000,
Peng2010}. We estimate that yield erosion has decreased the yield of
medium grain rice cultivars released by the CCRRF over the last 30+
years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1}, potentially
offsetting any gains made by the crop improvement program. However, it
is not clear what is responsible for this erosion. It is unlikely
biotic factors as pest and disease pressure are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. While efforts should be made to
identify the cause of erosion in this environment, this observation
highlights both the need for maintenance of plant breeding and the
difficulty of increasing genetic yield potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}. Without maintenance breeding (i.e., continuously
selecting cultivars well adapted to current conditions), CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields~\citep{espe2017}, and the RES experiences warmer
temperatures than many other rice growing areas in CA
\citep{espe2017}. These cultivars may perform well at the RES but not
widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield. The evidence here is consistent with the observation that the
genetic yield potential has not increased substantially in the last
30+ years (Fig. \ref{fig:first_yrs}).

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality~\citep{mckenzie1993} and
milling yield~\citep{mckenzie1994}, have been made over the study
period and are important. Also, crop duration has been reduced. For
example, during this study period, M-202 and M-206 have been the
predominate cultivars grown in CA, and while their yield potentials
were similar at the time of release (Fig. 5), M-206 matures about 6
days earlier than M-202 (Table \ref{tbl:ids}). Shorter duration is very important
in CA due to an already short growing season and water limitations
(shorter duration cultivars would presumably require less irrigation
water).

California state average yields have increased slowly
despite there being little to no gains in the breeding trials
(Fig. \ref{fig:state_yield}). The evidence here suggests mechanisms
other than increased genetic yield potential are responsible for these gains
(Fig. \ref{fig:compare_yrs}). Improvement in agronomic management,
including more efficient nutrient management~\citep{Lundy2012,
Lundy2015, linquist2009}, weed management~\citep{Saito2008,
Pittelkow2012, linquist2008}, and other factors may all be contributing to
increases in the state average yield. However, these gains are far less
than those experienced during the preceding era, and by their
nature agronomic gains cannot surpass the genetic yield potential
\citep{Fischer2015}. Judging from yields achieved at the RES
(Figs. \ref{fig:RES_year} \& \ref{fig:compare_yrs}), a
researcher-managed production system implementing best management
practices in more controlled situations compared to farmers' fields, state
average yields (approximately 10 Mg ha\tsups{-1} at the time of
writing; Fig. \ref{fig:state_yield}) may be approaching the current
genetic limit. Assuming that the level of management possible in small
experimental plots cannot be attained in farmers' fields and hence
experimental yields represent the attainable yield ceiling
\citep{Fischer2015}, agronomic gains in farmer's fields are nearing
their potential, consistent with physiological model-based estimates
of the attainable yield gap in CA rice production systems
\citep{espe16a}.

>>>>

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using 33 years of experimental
yield data from the California Cooperative Rice Research Foundation
Rice Experiment Station. Based on side-by-side comparisons of old and
new rice varieties which do not consider yield erosion, there was an
apparent increase in yield. However, the yields of older varieties
eroded at an estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1}
(90\% credible interval -4.4 to -53.3) after initial selection. Once
this effect was considered, the yield advantage of newer cultivars
over old was uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\%
credible interval -36.1 to 31.5). These results highlight first, the
importance of continuous crop improvement and deployment of new
cultivars simply to maintain existing yields, and second, to increase
the genetic yield potential, higher yield targets are
needed. Importantly, when breeding near the yield potential, despite
the limited yield gains, significant advances in improving quality and
reducing crop duration have been made.

@@

<<echo=FALSE>>=
library(rstan)
library(rstanarm)
library(lme4)
library(xtable)

@@

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

@
\caption{Average rice yields over the period \Sexpr{min(CA_data$Year)}
to \Sexpr{max(CA_data$Year)} for California, USA (US Dept. of
Agriculture National Agricultural Statistics Service, 2017).}
\label{fig:state_yield}
\end{figure}

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. For
comparison to estimates where yield erosion is not taken into account,
the model was used as described above except omitting the
$Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit to estimate
if there were differences between cultivars in the rate of yield
erosion via a random slope by cultivar (Model 3).

Data were processed and models fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval~\citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods, specifically a frequentist mixed-effects
model fit by maximum likelihood (see supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

@@

There was general agreement between the estimates from the Bayesian
and classical approaches for all models, with the estimates being
similar between approaches for similar models. The Bayesian estimates
are presented here due to the inherent flexibility of Bayesian
inference, though the classical estimates can be found in the
supplemental material. Comparing the cultivars' yield during each's
initial time (average 3 years) in the experimental trials just before
being released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1} (M-205)
(Fig \ref{fig:first_yrs}. Evaluating these initial yields over time
provides evidence of genetic yield improvements during the 33-year
study period. Individual exceptions to this were two cultivars (M-204
and M-205) that had higher yields than the others.

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3), hence the model excluding these
effects (Model 2) is presented here. The yields of each cultivar over
time were evaluated relative to when the cultivar was released using
the full model which accounts for both yield erosion and yield
improvements over time. The average decrease in yield over time (i.e.,
yield erosion) once a cultivar was selected was estimated at
\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI:
\Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year}; Table
\ref{tbl:coef}). The oldest cultivar in the trials
(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
is estimated to have lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline. M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed
solely to the erosion of M-202's yield over time (Figs
\ref{fig:first_yrs} \& \ref{fig:compare_yrs}). Further supporting this
estimate of yield erosion, both cultivars yielded similarly at their
respective year of release (Fig. \ref{fig:compare_yrs}), despite the
fact that M-202 was released 19 years prior to M-206
(Fig. \ref{fig:compare_yrs}; Table \ref{tbl:ids}).

@@

When yield erosion is not accounted for in the analysis (i.e.,
omitted from the model), there is estimated to be an increase per year
of approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}). However, the
full model which accounts for both the year of release and yield
erosion indicates that the estimated change in grain yield per year
in the CCRRF's breeding trials at the RES was \Sexpr{release_yr_coef}
kg ha\tsups{-1} year\tsups{-1}, with both small positive and negative
trends within the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}).

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated via the random
intercept term in the model). The marginal estimate for the
performance of a cultivar averaged over the uncertainty in the other
parameters in the model (i.e., the random effect) shows three
cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).

@@

In California, the cultivars with broad adaptation and adoption over
the 33 year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in these cultivars at any given time during the study
period~\citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
and M-402 which is a premium quality cultivar. Other cultivars such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) that does not account for yield erosion is
lower than the 37 to 60 kg ha\tsups{-1} year\tsups{-1} estimated
by~\citet{Omar2014} in side-by-side comparisons. However, when
accounting for yield erosion and looking acrossall cultivars, there is
no conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this cultivar is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or cooler
 temperatures reducing yields. Again, this is
unlikely to be the case in this situation as ~\citet{espe2017} found
no significant changes in growing season temperatures (either high,
low or average temps) the Sacramento Valley during the study period.
Thirdly, biotic and abiotic forces have previously been speculated as
the cause for yield erosion of the cultivar IR8 at
IRRI~\citep{Peng1998, Peng2000, Peng2010}. They argued that since the
cultivar's genetics remain stable from the time of selection but the
conditions the cultivar experiences over time are continuously
changing leading ot yield erosion~\citep{Peng1998, Peng2000,
Peng2010}. However, it is not clear what is responsible for the yield
erosion of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in
this analysis. It is unlikely biotic factors as pest and disease
pressure are low in this Mediterranean climate. Furthermore, these
fields are fallowed every other year and the residues are burned,
further lowering the potential for pest and disease pressure. While
efforts should be made to identify the cause of erosion in this
environment, this observation highlights both the need for maintenance
of plant breeding and the difficulty of increasing genetic yield
~\citep{espe2017} potential, especially in systems such as CA where yield is approaching
the physiological limit \citep{espe16a}. Without maintenance breeding
(i.e., continuously selecting cultivars well adapted to current
conditions), CA rice production systems are predicted to have lost an
estimated \Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain
yield over the study period.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields~\citep{espe2017}, and the RES experiences warmer
temperatures than many other rice growing areas in CA
\citep{espe2017}. Hence, cultivars which perform well at the RES might
not perform well across of CA. Since these stressful temperature
events are inconsistent across years, they influence yield stability (how consistent
yield is year to year). Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield.

@@

To be clear, our investigation here is narrowly focused on the
rough-rice/field grain yield. Other advancements due to plant
improvement, including increased quality~\citep{mckenzie1993} and
milling yield~\citep{mckenzie1994}, have been made over the study
period and are important. Maintaining high quality is imperative for
the breeding program and it is not uncommon in the breeding program
for higher yielding cultivars to be eliminated due to concerns about
quality. However, the goal of increasing yield may require the yield
performance of cultivars to be assessed earlier in the selection
process to avoid discarding these higher yielding entries without full
consideration. Another change in CA rice cultivars is that crop
duration has been reduced. For example, during this study period,
M-202 and M-206 have been the predominate cultivars grown in CA, and
while their yield potentials were similar at the time of release
(Fig. 5), M-206 matures about 6 days earlier than M-202 (Table
\ref{tbl:ids}). Shorter duration is very important in CA due to an
already short growing season and water limitations (shorter duration
cultivars would presumably require less irrigation water).

California state average yields have increased slowly during the study
period (Fig. \ref{fig:state_yield}) \citep{espe16a}. What are the
potential causes?  It is not due to the adoption of the broadly
adapted cultivars (M-202, M-206, and M-105) as there is no evidence of
increased yield potential among these (Fig. \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). While M-205 has increase yield potential, it
is not broadly adapted. Its adoption (although limited) in the
northern and warmer part of CA may contribute to some yield gains
reported at the state level. Other factors include improvement in
agronomic management, including more efficient nutrient
management~\citep{Lundy2012, Lundy2015, linquist2009}, and weed
management~\citep{Saito2008, Pittelkow2012, linquist2008}. However,
these state wide yield gains are far less than those experienced
during the preceding era, and by their nature agronomic gains cannot
surpass the genetic yield potential \citep{Fischer2015}. Judging from
yields achieved at the RES (Figs. \ref{fig:RES_year} \&
\ref{fig:compare_yrs}), a researcher-managed production system
implementing best management practices in more controlled situations
compared to farmers' fields, state average yields (approximately 10 Mg
ha\tsups{-1} at the time of writing; Fig. \ref{fig:state_yield}) may
be approaching the current genetic limit. Assuming that the level of
management possible in small experimental plots cannot be attained in
farmers' fields and hence experimental yields represent the attainable
yield ceiling \citep{Fischer2015}, agronomic gains in farmer's fields
are nearing their potential, consistent with physiological model-based
estimates of the attainable yield gap in CA rice production systems
\citep{espe16a}.
